(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}

### 2015-04-28 ms/ms.Rnw
Added VSURF for spatially-averaged climate

@@

\begin{table}[ht]
   \centering
   \topcaption{16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "HAU"], 3)}	& 
    \Sexpr{round(focClim$Longitude[focClim$ID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "HAU"]}   \\
	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "CTC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CTC"]}   \\
	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$ID == "CUR"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CUR"]}   \\
	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "GRP"], 3)} &
    \Sexpr{round(focClim$Longitude[focClim$ID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "GRP"]}   \\
	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WWC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WWC"]}   \\
	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "MIL"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MIL"]}   \\
	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "WFM"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WFM"]}   \\
	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$ID == "NMT"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "NMT"]}   \\
	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "PRD"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "PRD"]}   \\
	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RWD"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RWD"]}   \\
	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WNA"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WNA"]}   \\
	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "RBW"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RBW"]}   \\
	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "MYU"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MYU"]}   \\
	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "LIJ"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "LIJ"]}   \\
	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "DPC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "DPC"]}   \\
	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RCC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RCC"]}   \\
	\bottomrule
   \end{tabular}
\end{table}

@@

  traitPC <- prcomp(popMeans, center = T, scale. = T)
  focClim$PC1 <- traitPC$x[, "PC1"]

@@

As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude. We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We did this by looking for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations. We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table~\ref{table:TableS_}). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

@@

  # Variable selection using random forests (VSURF)
  linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
  set.seed(69647)
  vsurfPC1 <- VSURF.parallel(linmod.PC1, data = focClim[, c("PC1", climVars)], 
    ncores = 3)

@@

  # Overlap between latitude predictors and phenotype predictors
  imp2 <- climVars[vsurfPC1$varselect.pred]
  nImpPC1 <- length(imp2)

@@

\begin{table}[ht]
   \centering
   \topcaption{Climatic variables}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Climate variable & Abbreviation \\
      \midrule
  % climateWNA VARIABLES:
	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
	NFFD	& number of frost-free days                           \\
	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
	Eref	& Hargreaves reference evaporation (mm) \\
	CMD	  & Hargreaves climativ moisture deficit (mm) \\
	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
	RH    &	mean annual relative humidity \\
  % BIOCLIM VARIABLES:
  bio1	&	Annual Mean Temperature	\\
	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
	bio4	&	Temperature Seasonality (standard deviation *100)	\\
	bio5	&	Max Temperature of Warmest Month	\\
	bio6	&	Min Temperature of Coldest Month	\\
	bio7	&	Temperature Annual Range (bio5-bio6)	\\
	bio8	&	Mean Temperature of Wettest Quarter	\\
	bio9	&	Mean Temperature of Driest Quarter	\\
	bio10	&	Mean Temperature of Warmest Quarter	\\
	bio11	&	Mean Temperature of Coldest Quarter	\\
	bio12	&	Annual Precipitation	\\
	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
	bio16	&	Precipitation of Wettest Quarter	\\
	bio17	&	Precipitation of Driest Quarter	\\
	bio18	&	Precipitation of Warmest Quarter	\\
	bio19	&	Precipitation of Coldest Quarter	\\
	    \bottomrule
   \end{tabular}
\end{table}

>>>>

# Read in spatially-averaged (10 km buffer) for 16 focal populations
load(paste(pathDat, "/focClimSA.RData", sep = ""))

@@

\begin{table}[ht]
   \centering
   \topcaption{16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Hauser Creek"], 3)}	& 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Hauser Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Hauser Creek"]}   \\
	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Cottonwood Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cottonwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cottonwood Creek"]}   \\
	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$PopID == "Cuyamaca Rancho"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cuyamaca Rancho"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cuyamaca Rancho"]}   \\
	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Grade Road Palomar"], 3)} &
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Grade Road Palomar"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Grade Road Palomar"]}   \\
	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Whitewater Canyon"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Whitewater Canyon"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Whitewater Canyon"]}   \\
	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Mill Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Mill Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Mill Creek"]}   \\
	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "West Fork Mojave River"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "West Fork Mojave River"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "West Fork Mojave River"]}   \\
	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$PopID == "North Fork Middle Tule"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "North Fork Middle Tule"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "North Fork Middle Tule"]}   \\
	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Paradise Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Paradise Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Paradise Creek"]}   \\
	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Redwood Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Redwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Redwood Creek"]}   \\
	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Wawona"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Wawona"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Wawona"]}   \\
	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rainbow Pool"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rainbow Pool"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rainbow Pool"]}   \\
	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Middle Yuba"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Middle Yuba"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Middle Yuba"]}   \\
	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Little Jameson"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Little Jameson"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Little Jameson"]}   \\
	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Deep Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Deep Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Deep Creek"]}   \\
	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rock Creek"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rock Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rock Creek"]}   \\
	\bottomrule
   \end{tabular}
\end{table}

@@

  traitPC <- prcomp(popMeans, center = T, scale. = T)
  focClim$PC1 <- traitPC$x[, "PC1"]
  focClimSA$PC1 <- traitPC$x[, "PC1"]

@@

As with previous studies in \textit{Mimulus} and many other species [CITES, Huey_etal, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude. We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We did this by looking for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations. We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included \Sexpr{length(climVars) / 2} cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

@@

  # Variable selection using random forests (VSURF)
  linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
  set.seed(69647)
  vsurfPC1 <- VSURF.parallel(linmod.PC1, data = focClim[, c("PC1", climVars)], 
    ncores = 3)
  vsurfPC1_SA <- VSURF.parallel(linmod.PC1, data = focClimSA[, c("PC1", climVars)], 
    ncores = 3)

@@

  # Overlap between latitude predictors and phenotype predictors
#### NEED TO COMPARE RESULTS FROM POINT DATA AND SPATIALLY AVERAGED CLIMATE!
  imp2 <- climVars[vsurfPC1$varselect.pred]
  # imp2 <- climVars[vsurfPC1_SA$varselect.pred]
  nImpPC1 <- length(imp2)

@@

\begin{table}[ht]
   \centering
   \topcaption{Climatic variables}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Climate variable & Abbreviation \\
      \midrule
  % climateWNA VARIABLES:
	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
	NFFD	& number of frost-free days                           \\
	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
	Eref	& Hargreaves reference evaporation (mm) \\
	CMD	  & Hargreaves climativ moisture deficit (mm) \\
	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
	RH    &	mean annual relative humidity \\
  % BIOCLIM VARIABLES:
  bio1	&	Annual Mean Temperature	\\
	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
	bio4	&	Temperature Seasonality (standard deviation *100)	\\
	bio5	&	Max Temperature of Warmest Month	\\
	bio6	&	Min Temperature of Coldest Month	\\
	bio7	&	Temperature Annual Range (bio5-bio6)	\\
	bio8	&	Mean Temperature of Wettest Quarter	\\
	bio9	&	Mean Temperature of Driest Quarter	\\
	bio10	&	Mean Temperature of Warmest Quarter	\\
	bio11	&	Mean Temperature of Coldest Quarter	\\
	bio12	&	Annual Precipitation	\\
	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
	bio16	&	Precipitation of Wettest Quarter	\\
	bio17	&	Precipitation of Driest Quarter	\\
	bio18	&	Precipitation of Warmest Quarter	\\
	bio19	&	Precipitation of Coldest Quarter	\\
	    \bottomrule
   \end{tabular}
   \label{table:TableS_ClimVars}
\end{table}
